Dynamic Stochastic Approximation Monte Carlo

### 2018-06-25 papers/histogram/sad.tex
simplify sad algorithm at extreme energies (MUST RERUN ALL SAD DATA!)

\subsubsection{High and Low energy racheting}
Since SAD explores all energy states it is imperative that the method
have a reasonable idea about where the minimum important energy and
maximum entropy state are located in energy space. Since these are not
user input parameters such as they are for methods like WL-TMMC, the
simulation is responsible for determining and updating these energy
locations. In order to prevent, the method from experiencing errors
near the energy extrema, we employ a high energy $\epsilon_H$ and a low
energy $\epsilon_L$ to define the range over which the energy histogram
is made flat. $\epsilon_H$ is the highest ever value that the maximum
entropy state has taken while $\epsilon_L$ is the lowest ever value
that the minimum important energy has taken. SAD uses these two
energies to control the weights $w(E)$ above $\epsilon_H$ and
$\epsilon_L$. If the energy $E$ is greater than $\epsilon_H$ then the
weights can be calculated as ~\jpsays{equation formatting problems!}
\begin{equation}
\begin{aligned}
\ln w(E) = \max(\ln w(E) + \gamma, \\
    \ln w(\epsilon_H) + \log(e^{w(E)-\ln w(\epsilon_H)}))
\end{aligned}
\end{equation}
If the energy $E$ is less than $\epsilon_L$ then the weights can be
calculated as
\begin{equation}
\begin{aligned}
\ln w(E) = \max(\ln w(E) + \gamma, \ln w(\epsilon_L) \\
    + \frac{\epsilon_L - E}{T_\text{min}}
    + \log(\gamma + e^{w(E) -\ln w(\epsilon_L)
    + \frac{\epsilon_L -E}{T_\text{min}}}))
\end{aligned}
\end{equation}
Sad sets the maximum entropy state equal to the energy if either of the
following is true.
\begin{align}
\ln w(E) > \ln w(S_\text{max}) \; \text{or} \; H(S_\text{max}) = 0
\end{align}
Thus if the method is beyond the maximum entropy state or has collected no
statistics at the maximum entropy state then it is set equal to the energy.
Similarly, Sad sets the minimum important energy equal to the energy if
either of the following is true.
~\jpsays{equation formatting problems!}
\begin{align}
\ln w(E) > \ln w(E_\text{min}) - \frac{E_\text{min}-E}{T_\text{min}}
\; \text{or} \; H(E_\text{min}) = 0
\end{align}
Thus if the method is beyond the minimum important energy or has collected no
statistics at the minimum important energy then it is set equal to the energy.
Together these two energies serve as a kind of ratcheting effect to
determine the energy landscape boundaries. Outside the range set by the
weights $w(E)$, the maximum entropy state is set to the energy $E$. The
minimum important energy is also set to the energy when outside the
calculated range.

>>>>

\subsubsection{High and Low energy racheting}
Since SAD explores all energy states it is imperative that the method
have a reasonable idea about where the minimum important energy and
maximum entropy state are located in energy space. Since these are not
user input parameters such as they are for methods like WL-TMMC, the
simulation is responsible for determining and updating these energy
locations. In order to prevent, the method from experiencing errors
near the energy extrema, we employ a high energy $\epsilon_H$ and a low
energy $\epsilon_L$ to define the range over which the energy histogram
is made flat. $\epsilon_H$ is the highest ever value that the maximum
entropy state has taken while $\epsilon_L$ is the lowest ever value
that the minimum important energy has taken. SAD uses these two
energies to control the weights $w(E)$ above $\epsilon_H$ and
$\epsilon_L$. If the energy $E$ is greater than $\epsilon_H$ then the
weights can be calculated as ~\jpsays{equation formatting problems!}
\begin{equation}
\begin{aligned}
\ln w(E) = \max(\ln w(E) + \gamma, \\
    \ln w(\epsilon_H) + \log(\gamma + e^{\ln w(E)-\ln w(\epsilon_H)}))
\end{aligned}
\end{equation}
If the energy $E$ is less than $\epsilon_L$ then the weights can be
calculated as
\begin{equation}
\begin{aligned}
\ln w(E) = \max(\ln w(E) + \gamma, \ln w(\epsilon_L) \\
    + \frac{\epsilon_L - E}{T_\text{min}}
    + \log(\gamma + e^{\ln w(E) -\ln w(\epsilon_L)
    + \frac{\epsilon_L -E}{T_\text{min}}}))
\end{aligned}
\end{equation}
Sad sets the maximum entropy state equal to the energy if either of the
following is true.
\begin{align}
\ln w(E) > \ln w(S_\text{max}) \; \text{or} \; H(S_\text{max}) = 0
\end{align}
Thus if the method is beyond the maximum entropy state or has collected no
statistics at the maximum entropy state then it is set equal to the energy.
Similarly, Sad sets the minimum important energy equal to the energy if
either of the following is true.
~\jpsays{equation formatting problems!}
\begin{align}
\ln w(E) > \ln w(E_\text{min}) - \frac{E_\text{min}-E}{T_\text{min}}
\; \text{or} \; H(E_\text{min}) = 0
\end{align}
Thus if the method is beyond the minimum important energy or has collected no
statistics at the minimum important energy then it is set equal to the energy.
Together these two energies serve as a kind of ratcheting effect to
determine the energy landscape boundaries. Outside the range set by the
weights $w(E)$, the maximum entropy state is set to the energy $E$. The
minimum important energy is also set to the energy when outside the
calculated range.

\paragraph{Accepting moves above and below}
We do \emph{not} use the weights to determine move acceptance outside
the middle range.  FIXME WRITE THIS SECTION.

\paragraph{Above $\epsilon_H$}
Each energy $E>\epsilon_H$, should be visited at a rate that
is proportional to the density of states at that energy, as in a
simulation with constant weights.  Thus the weight at each energy
needs to be proportional to the number of times that weight has been
visited.  In order to ensure that $w(E) \propto D(E)$, the change in
weight with each visit should be equal to the change in weight at
the border energy $E=\epsilon_H$.
\begin{align}
  \Delta w(E) &= \Delta w(\epsilon_H) \\
  w_{\text{new}}(E)-w_{\text{old}}(E)
    &= e^\gamma w(\epsilon_H) - w(\epsilon_H) \\
    &= (e^\gamma-1) w(\epsilon_H)
\end{align}
This ensures that although we visit each energy above $\epsilon_H$ only
in proportion to its density of states, the weights at these energies
will approach the entropy as the simulation progresses.

\paragraph{Below $\epsilon_L$}
For energies $E<\epsilon_L$, we will visit each state in proportion to
the Boltzmann distribution at temperature $T_{\min}$.  This ensures that
we have adequate statistics to describe the system at temperature
$T_{\min}$ while not wasting time simulating extremely low energies with
correspondingly low entropy.  This means that we need to shift the
weights at lower energies by a smaller amount in proportion to the
Boltzmann factor.
\begin{align}
  \Delta w(E) &= \Delta w(\epsilon_L)e^{(E-\epsilon_L)/T_{\min}} \\
  w_{\text{new}}(E)-w_{\text{old}}(E)
    %&= \left(e^\gamma w(\epsilon_L) - w(\epsilon_L)\right)e^{(E-\epsilon_L)/T_{\min}} \\
    &= (e^\gamma-1) w(\epsilon_L)e^{(E-\epsilon_L)/T_{\min}}
\end{align}
We note that when updating the natural logarithms of the weights, care
is required in order to ensure that overflow does not occur.
